# Objects, Functions and The Pipe Operator %>%

Everything in R is an object

Make an object and use str to look at its type

```{r}
x <- 10
str(x)
```

Functions are things that operate on objects.
For example, we can use sum to add numbers.

```{r}
sum(1,2,3)
```


Functions are also objects

```{r}
str(sum)
```

You can make your own functions

```{r}
jamessum <- function(a,b) {
  return(a+b+1)
}

jamessum(1,2)
```

Functions are written as verb(object)

When you see functions written down it looks like verb(noun)
Pick up the cat ===> pickup(cat)

This makes sense when you are only doing one thing to an object or a group of objects

But often we are doing multiple things to to the cat.
When we're talking about this we might say:

Get the cat
 - then pick it up
 - then put it on your lap
 - then stoke it.

The pipe operator %>% is a useful thing that let us do this


cat %>%
  pick_up() %>%
  put_on_lap() %>%
  stroke()


Whereas in our normal syntax land we would have to do this...

stroke(put_on_lap(pick_up(cat)))

...which you'll probably agree is a lot more confusing
It is backwards compared to the way you understand it as a human.


The pipe operator let's us chain together operations on the same object.

The tidyverse package functions are written in a consistent way that makes it easy to chain things together.

Install and load tidyverse

```{r}
install.packages('tidyverse')
library(tidyverse)
```


We'll look at a real example, but first let's just explain exactly what that operator is doing.... 

Remember our sum operator, we can use it to add up numbers

```{r}
sum(1,2,3)
```

Those things that we passed to the function are called arguments.
The pipe operator just lets us insert the first of those arguments from outside the function itself

```{r}
1 %>% sum(2,3)
```


That's it really. The pipe just makes it easy to chain stuff together

```{r}
x %>% 
  sum(10) %>%
  max(1,2,4) %>%
  sqrt()
```

