---
title: "Rolling Correlation of SPX Constituents"
output: html_notebook
---

Given a database of stock prices, how might we calculate rolling correlations between constituents of the S&P 500 index?

For problems like this, the `tidyverse` really shines. There are a number of ways to solve this problem ... read on for our solution, and let us know if you'd approach it differently.

First, we load some packages and some data that we extracted earlier. `spxprices.RData` contains a dataframe, `prices_df`, of S&P500 constituents and their daily prices. It also contains a boolean flag `inSPX` which defines whether the stock was in the index that month. (we prepared this upstream using BigQuery).

```{r, warning = FALSE}
library(tidyverse)
library(glue)
library(here)

load(here::here("data", "spxprices.RData"))
prices_df %>%
  head()
```

We'd like to be able to calculate rolling average pairwise correlations between all the stocks (or subsets of them) in as tidy a way possible.

That is, we would calculate the rolling pairwise correlation between all the stock combinations in the index and then take the mean of all those.

Our general aproach was to:

- calculating returns for each ticker 
- create a long dataframe of all the combinations that were in the index for each day by doing a full join of the data on itself
- remove instances where we had the same stock twice (corresponding to the diagonal of the correlation matrix)
- remove instances where we have the same stock, but in different columns, eg we only want one of APPL-GOOG and GOOG-APPL (this is equivalent to removing the upper or lower triangle of the correlation matrix)
- use `slider::slide2_dbl` to do the rolling correlation calculation 
- group by date and take the mean

We run into trouble trying to do the full join on all the data - the resulting object is about 15GB in size. 

We'll solve that later, but for now we'll just take a sample of the price data and work with that.

Here's the code for doing that:


```{r}
# calculate returns to each stock
df <- prices_df %>%
  # filter(ticker %in% c("A", "AAL", "AAP")) %>%
  filter(date >= "2019-01-01", date < "2020-01-01") %>%  # try to do it on a single year
  group_by(ticker) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(return = close / dplyr::lag(close) - 1) %>%
  select(date, ticker, return, inSPX)

# combinations by date, where both tickers are index members
pairwise_combos <- df %>%
  full_join(df, by = "date") %>%
  filter(ticker.x != ticker.y, inSPX.x == TRUE, inSPX.y == TRUE) %>%
  select(date, ticker.x, ticker.y, return.x, return.y)

# remove duplicate pairs (eg A-AAL, AAL-A)
pairwise_combos <- pairwise_combos %>%
  ungroup() %>%  # this is a gotcha - if we don't ungroup our variables, the grouping variable will be added back and thwart attempts to filter distinct cases.
  mutate(tickers = ifelse(ticker.x < ticker.y, glue("{ticker.x}, {ticker.y}"), glue("{ticker.y}, {ticker.x}"))) %>%
  distinct(date, tickers, .keep_all = TRUE) 

period <- 60

pairwise_corrs <- pairwise_combos %>%
  group_by(tickers) %>%
  arrange(date, .by_group = TRUE) %>%
  mutate(rollingcor = slider::slide2_dbl(.x = return.x, .y = return.y, .f = ~cor(.x, .y), .before = period, .complete = TRUE)) %>%
  select(date, tickers, rollingcor)

mean_pw_cors <- pairwise_corrs %>%
  group_by(date) %>%
  summarise(mean_pw_corr = mean(rollingcor, na.rm = TRUE))

mean_pw_cors %>%
  na.omit() %>%
  ggplot(aes(x = date, y = mean_pw_corr)) +
    geom_line() +
    labs(
      x = "Date",
      y = "Mean Pairwise Correlation",
      title = "Rolling Mean Pairwise Correlation",
      subtitle = "SPX Constituents"
    )
  
```

Blog:
Illustrate problem doing it in base R. 64 million rows for a single year, gets whittled down to 32 million


Some plots.....

```{r}
pairwise_corrs %>%
  na.omit() %>%
  filter(tickers %in% c("A, AAL", "A, AAP", "AAL, AAP")) %>%
  ggplot(aes(x = date, y = rollingcor, colour = tickers)) +
    geom_line() +
    labs(
      x = "Date", 
      y = glue("{period}-day rolling correlation"), 
      title = "Rolling correlation of SPX constituents", 
      colour = "Tickers"
    ) +
    theme_bw()

p <- pairwise_corrs %>%
  na.omit() %>%
  filter(tickers %in% c("A, AAL", "A, AAP", "AAL, AAP")) %>%
  ggplot(aes(x = date, y = rollingcor)) +
    geom_line() +
    facet_wrap(~tickers, ncol = 1) +
    labs(
      x = "Date", 
      y = glue("{period}-day rolling correlation"), 
      title = "Rolling correlation of SPX constituents" 
    ) +
    facet_wrap(~tickers, ncol = 1) +
    theme_bw()

p
```


```{r}
library(showtext)
font_add_google("Gochi Hand", "gochi")

showtext_auto()
p + theme_bw(base_family =  "gochi")
```

