---
title: "dplyr for traders"
output: github_document 
always_allow_html: true
---

We're going to show how a quant trader can manipulate stock price data using the dplyr package.

```{r}
if (!require("pacman")) install.packages("pacman")
pacman::p_load(tidyverse, here)
```

First, load some price data. 

`energystockprices.RDS` contains a data frame of daily price observations for 3 energy stocks.

```{r}
prices <- readRDS(here::here('data','energystockprices.RDS'))
prices
```

We've organised our data so that

- Every column is variable.
- Every row is an observation.

In this data set:

- We have 13,314 rows in our data frame.
- Each row represents a daily price *observation* for a given stock.
- For each observation measure the open, high, low and close prices, and the volume traded.

This is a very helpful way to stucture your price data. We'll see how we can use the dplyr package to manipulate price data for quant analysis.

# The main dplyr verbs

There are 6 main functions to master in `dplyr`. 

- `filter()` picks outs observations (rows) by some filter criteria
- `arrange()` reorders the observations (rows)
- `select()` picks out the variables (columns)
- `mutate()` creates new variables (columns) by applying transformations to existing variables
- `summarise()` allows you to group and summarise data - reducing the data into a grouped summary with fewer rows.

Finally, the `group_by()` causes the verbs above to act on a group at a time, rather than the whole dataset.

We'll go through them one by one.

## Filter

### Filtering rows for a single stock with `filter()` and ==

```{r}
prices %>%
  filter(ticker == 'XOM')
```

`filter()` returns all observations (rows) which satisfy the criteria within the `filter()` function.

### Filtering rows for multiple stocks with `filter()` and %in%

```{r}
prices %>%
  filter(ticker %in% c('XOM','CVX'))
```

## Arrange

`arrange()` reorders the rows in your data by one or more criteria.

## Reordering rows in order of volume traded ascending with `arrange()`

```{r}
prices %>%
  arrange(desc(volume))
```

## Reordering rows in order of date descending, then volume ascending with `arrange()`

```{r}
prices %>%
  arrange(desc(date), volume)
```

# Select

## Selecting columns with `select()`

Here we use `select()` to pull out the ticker, date, close columns.

```{r}
prices %>%
  select(ticker, date, close)
```

## Chaining transformations together: Filtering, electing and reordering. 

Now we'll see how we can chain transformations together. We're going to filter out a single stock, then select the date and close price, then reorder by date descending.

```{r}
prices %>%
  filter(ticker == 'XOM') %>%
  select(date, close) %>%
  arrange(desc(date))
```

When debugging a `dplyr` pipeline, it can be helpful to select each transformation one by one, and press CTRL-Enter in Rstudio to just run the selected transformation.

# Mutate

Now we get onto slightly more exciting stuff. 

`mutate()` lets us add variables (columns) to our data. These are usually derived from existing columns.

## Add new variable (column) for log volume with `mutate()`

Here we use mutate to add a column called `logvolume` to our data. It is calculated as the log of the `volume` variable.

```{r}
prices %>%
  mutate(logvolume = log(volume))
```

## Add new variable (column) for intraday returns with `mutate()`

Now we add a column for the intraday (open to close) simple returns of the stock each day.

```{r}
prices %>%
  mutate(intradayreturns = close / open - 1)
```

# Summarise

`Summarise()` summarises all our observations to a single row.

## Calculate the mean volume traded for all observations using `summarise()`

Let's use it to calculate the mean volume traded over our data set. 

```{r}
prices %>%
  summarise(meanvolume = mean(volume),
            count = n())
```

## Use `summarise()` with `group_by()` to calculate mean traded volume for each stock

That's not always very useful. 

Usually you want to group by a variable, and then summarise that grouped data.

The `group_by()` function tells the dplyr verbs to operate on each group one at a time. 

If we group by ticker, then call summarise, then `dplyr` will preform the summary calculations separately for each ticker. We will get a row for each ticker.

```{r}
prices %>% 
  group_by(ticker) %>%
  summarise(meanvolume = mean(volume))
```

## Use `summarise()` with multiple `group_by` variables, to calculate the mean traded volume for each stock for each year

```{r}
library(lubridate)
prices %>%
  mutate(year = year(date)) %>%
  group_by(ticker, year) %>%
  summarise(meanvolume = mean(volume),
            obscount = n())
```


# `group_by` with `mutate()`

We can also use `group_by` with `mutate()` to calculate new variables which are calculated separately for a given variable (or set of variables)

One way you'll use this nearly everytime you do any quant analysis is to calculate periodic returns. 

## Using `group_by` with `mutate()` and `lag()` to calculate daily close-to-close returns

```{r}
prices %>%
  group_by(ticker) %>%
  arrange(date) %>%
  mutate(c2creturns = close / lag(close) - 1)
```

# Summary

Arrange your data so:

- Every column is variable
- Every row is an observation

You can then easily use `dplyr` to manipulate that data very efficiently. 

There are 6 main functions to master in `dplyr`. 

- `filter()` picks outs observations (rows) by some filter criteria
- `arrange()` reorders the observations (rows)
- `select()` picks out the variables (columns)
- `mutate()` creates new variables (columns) by applying transformations to existing variables
- `summarise()` allows you to group and summarise data - reducing the data into a grouped summary with fewer rows.

The `group_by()` causes the verbs above to act on a group at a time, rather than the whole dataset.
