---
title: "Tidy JSON Extraction"
output: html_notebook
---

Here are some tools and recipes for working with JSON data in the tidyverse. 

```{r}
library(tidyverse)
library(here)
```

## load json as nested named lists

This data has been converted from raw JSON to nested named lists using `jsonlite::fromJSON`

```{r}
strikes <- readRDS(here::here("data", "SPYstrikes.RDS"))
```

## look inside

```{r}
str(strikes, max.level = 1)  
```
This tells us we have a component named "data". Let's look at that a little more closely:

```{r}
str(strikes$data, max.level = 1, list.len = 10)
```

This suggests we have homogenous lists of 40 elements each (an assumption we'll check shortly). 

Let's look at one of those lists:

```{r}
str(strikes$data[[1]])
```
All these elements look like they can be easily handled. For instance, I don't see any more deeply nested lists, weird missing values, or anything else that looks difficult. 

So now I'll pull out the interesting bit:
```{r}
strikes <- strikes[["data"]]
```

## How many observations do we have?
```{r}
length(strikes)
```
## Are all strike sublists identically named?

This is where we'll check that our sublists are indeed homogeneously named, as we assumed above:

```{r}
strikes %>%
  map(names) %>%  # this applies the base R function names to each sublist, and returns a list of lists with the output
  unique() %>%
  length() == 1
```

## Make a dataframe

We should also check the variable types are consistent as we need single types in each column of a dataframe (although R will warn if it is forced to coerce one type to another).

Here's an interesting thing. It uses a nested map to get the variable types for each element of each sublist. They're actually not identical according to this:

```{r}
strikes %>%
  map(.f = ~{map_chr(.x, .f = class)}) %>%
  unique() %>%
  length()
```
After we make our dataframe, we should check that the variable types are as expected. 

Now, to that dataframe...

`purrr::flatten` removes one level of hierarchy from a list (`unlist` removes them all). Here, `flatten` is applied to each sub-list in `strikes` via `purrr:map_df`. 

We use the variant `flatten_df` which returns each sublist as a dataframe, which makes it compatible with `purrr::map_df`,which requires a function that returns a dataframe.

```{r}
strikes_df <- strikes %>%
  map_df(flatten_df)

strikes_df %>%
  head(10)
```

## Some other interesting things

### Get vector of column names
```{r}
strikes %>%
  map(names) %>%
  unique() %>%
  unlist()
```

### Check that all elements have the same ticker

```{r}
strikes %>%
  map_chr("ticker") %>%  # this makes a character vector of list elements "ticker"
  unique()
```

