---
title: "Functional programming in R"
output: html_notebook
---


# Introduction to Functional Programming

As a data scientist or quant researcher, functional programming is a thing you should care about. Why?

- problem solving centred on functions is a good fit for data analysis
- it matches the way we tend to think about the questions we ask of data
- it tends to improve the readability of code (for instance by replacing for loops with a single line of code that makes the intent obvious to the reader)

In R, functions behave like other data structures. That means you can assign them to a variable, create them inside other functions, return them as a result of a function, add them to lists, etc. 

```{r}
# function that returns a function
power <- function(n) {
  
  f = function(x) { x^n }
  
  return(f)
}

# make a function from our power function
cube <- power(3)

# use cube as a function
cube(1:3)

# assign it to a variable
a <- cube
print(cube)
a(4)

# add it to a list
my_stuff <- list(
  vec = 1:5,
  func = cube
)

my_stuff
```
That information by itself isn't that interesting...but the implications certainly are.

One of these implications is that you can take a function that solves a problem for a single input and then generalise it to handle any number of inputs. This is what base R's `apply` functions do:

```{r}
# get the median of each vector in a list:
vecs <- list(
  c(1:10),
  c(1:30),
  c(1:50)
)

meds <- lapply(vecs, FUN = median)
```

I mentioned that functionals are a neat and concise alternative to for loops. The `lapply` function above replaces this for loop:

```{r}
meds <- list()
for(i in seq_along(vecs)) {
  
  meds[[i]] <- median(vecs[[i]])

}

meds
```

...
One use case is an alternative to for looops - makes code much more readable and understandable (concise, plus when you see a functional, you know what it's used for, since each functional is rather specific).


## Some interesting things that functional programming enables:
Enables:

- rolling/sliding window calculations
- mapping functions over columns/variables

Functionals have a consistent interface - easier and more intuitive than the variants of apply.


## Syntax
~ creates a formula eg lm(mpg ~ wt)
Formulas prevent immediate evaluation of symbols. eg 

x <- ~f(a+b) can be executed without f, a or b being defined!

Why is that useful?

map needs a function, but it also accepts a formula, character/numeric vector or list which are converted to functions.

~ . can be used as an anonymous function call, in this case function(x) x

tilde-dot (`~.`) is shorthand for anonymous functions (`.p` and `.f` in the `modify_if` example). eg
```R
function(x) {
	x*2
}
```
becomes 

```R
~{. + 10}
```

Syntax:
In this case, `~` indicates the start of an anonymous function. The argument to the function can be referred to as `.` (sometimes you see this written `.x`). The `~.` function argument is *always* `.` or `.x`. 

Note that ~ also indicates a function eg lm(y ~ x). If we did map(z, ~lm(y ~ x, data = .)), R would interpret lm(...) as an anonymous function, with a formula inside. This syntax allows tidyverse functions to construct anonymous functions on the fly.

Note this aplies in the tidyverse... outside of that ~ in R refers to a formula, and . refers to all colums not used, eg lm(x ~ .,)

In the tidyverse, ~ turns a formula into a function.
The tilde notation is basically turning a formula into a function, which is most times easier to read. Each option can be turned into a named function, which works as follows. Ideally, the named function reduces the complexity of the underlying function to one argument (the one which should be looped over) and in this case the function can be called like all simple functions in map without further arguments/notations.

https://stackoverflow.com/questions/56621051/in-map-when-is-it-necessary-to-use-a-tilde-and-a-period-and

More correctly, ~ in tidyverse is a special case of the formula:
The tilde operator in R is a general operator that creates a formula object.
However, the usage in your specific pieces of code is a special case of that: purrr co-opts formulas to implement an anonymous function notation.
https://stackoverflow.com/questions/56650810/what-does-the-tilde-mean-in-this-context-of-r-code

## Simple examples
Something from purr
Something from slider


## Then go deep on purrr, slider

purrr: anytime you need to "loop through columns" consider purrr instead