---
title: "Functional programming in R"
output: html_notebook
---

Make a tutorial with finance examples. 

Then pull out each bit into a bite sized recipe.


```{r}
library(tidyverse)
```


## Syntaxes



## purrr

`purrr` provides tidy apply functions in the form of `map` and list manipulation functions. 

Focuses on using lists, while dplyr focuses on the dataframe. 

A map function applies the same function to every element of an object, such as each element of a list. Base R `apply` is an example of a map function. 


Functionals have a consistent interface - easier and more intuitive than the variants of apply.

### purrr::map
23 variants - but they're easy to remember

map takes a vector and a function and calls the function once for each element of the vector, returning the results as a list (it maps one vector (the input) to another (the output, represented as a list):

```{r}
square <- function(x) {
  x^3
}

map(1:3, square)
```

This is the same as doing 

```{r}
result <- list()
x <- 1:3
for(i in seq_along(x)) {
  result[[i]] <- square(x[[i]])
}

result
```
purrr::map is not only much mroe concise than the for loop version, but it's also written in C so is quite performant. 

I mentioned above that there are 18 variants of `map`. `map` is the most general, as it returns a list. The following variants work exactly the same way, but they **return** different data structures:

- `map_dbl`: a numeric vector
- `map_chr`: a character vector
- `map_lgl`: a boolean vector
- `map_int`: an integer
- `map_df`: dataframe

The first input to each map function is the object you want to iterate over. It must be a vector (iterate over elements), a list (iterate over elements), or a dataframe (iterate over columns). (lists and dataframes are lists of (column) vectors in R (is this true??))

Show what happens if you use say map_chr with with a function that doesn't return a character. Error, wrong type.

This is a common gotcha: writing a custom function that doesn't return the correct type for your map variant. df example.

Also note that the map functions return an output of the same length as the input - if you're expecting something else, then map isn't the right tool. If you use a function that returns more than a single element, you'll get an error.

#### Anonymous functions and a shortcut

We can define an anonymous function that we pass to the map function:

```{r}
map(df, function(x) mean(x)/sd(x))
```

Which is nice, but somewhat cumbersome. 

purrr supports the tilde-dot (`~.`) notation which we can use as shorthand for the above function:

```{r}
map(df, ~mean(.x)/sd(.x))
```

You might recognise `~` as representing a forumula in R. purrr translates such a formula into a function. `.x` replaces the function argument `x`. In single argument functions, you can write `.` in place of `.x` if you wish:

```{r}
map(df, ~mean(.)sd(.))
```

If your use case doesn't call for passing an argument, that's fine too, just use `~`:
```{r}
map(1:5, ~rnorm(10, 0, 1))
```

What's the benefit of using `~.` rather than writing and passing a separate function? It's really just a matter of style and readability: when you're function is short, your code will tend to be more concise and readable if you use `~.`. Consider also what's *implied* by the use of a separate function - that the function is used elsewhere in the code.

`map` also enables efficient extraction of elements of deeply nested lists. If instead of a function, you pass a name or index, `map` will return the corresponding element in each list. JSON parsing is a good use of this:

```{r}

```


`...` gets passed along as an optional argument after the function name"

```{r}
map_dbl(x, mean, na.rm = TRUE)
```


When the input vector element are to be second arguments to a function, use an anonymous function and .x:

```{r}

```


```{r}
map(1:3, rnorm(2))  # doesn't work because `rnorm` is evaluated once and the result passed back to map resulting in a pluck.
map(1:3, ~rnorm(2))  # works because `rnorm` is converted to an anonymous function and is thus applied as a functional argument to map
```


Mean of every column in a dataframe:

```{r}
df <- do.call(rbind, map(1:5, ~rnorm(5))) %>% as.data.frame()
map_dbl(df, ~mean(.x))
map_dbl(df, mean)

```

Another way to make a dataframe of random numbers (bind_rows needs named lists):

```{r}
map(1:5, ~rnorm(5)) %>%
  setNames(paste0("var_", c(1:5))) %>%
  bind_rows()
```
https://stackoverflow.com/questions/44834446/what-is-meaning-of-first-tilde-in-purrrmap





#### Other variants

23 variants of map, but that's not as complex as it sounds. We know five from their output. The rest follow the same pattern:

- `modify` works like `map` but always returns the same data type as the input. 
- `map2` allows us to iterate over two inputs
- `imap` is used for iterating over an index
- `walk` iterates but returns nothing
- `pmap` iterates over any number of inputs

`modify` - why use `modify` but not `map_df`?

`modify_if` applies the function `.f` to elements specified by a second predicate function, `.p`: 
```{r}
modify_if(.x = list(1, 2, 3), .p = function(x) x > 2, .f = function(x) x*2)
```


The `modify_if` example becomes 
```{r}
modify_if(.x = list(1, 2, 3), .p = ~{. > 2}, .f = ~{.*2})
```


approach to replacing for loops with functionals that fits really nicely with the purrr syntax.

### purrr:reduce
  reduce a vector to a single result
  
### predicates


### 



## pipe something into an argument other than first
mydata %>%
	lm(x ~ y, data = .)


## purr tricks
library(purrr)
map(x, 1) # get the first elements from a nested list

### get information about columns quickly
purrr::map_int(mtcars, n_distinct)
#>  mpg  cyl disp   hp drat   wt qsec   vs   am gear carb 
#>   25    3   27   22   22   29   30    2    2    3    6

# use map2 like python zip
win_rates <- seq(0.5, 0.65, by = 0.02)
initial_sizes <- c(0.0025, 0.005, 0.01)

df <- data.frame(
  win_rate = rep(win_rates, length(initial_sizes)), 
  initial_size = (
    map(initial_sizes, ~rep(., length(win_rates))) %>% 
      unlist()
  )
) 
df <- df%>%
  mutate(n_50 = map2_dbl(df$win_rate, df$initial_size, .f = n_50))

## map, map2 return a list
## map_dbl returns a numeric vector the same length as the input

```{r}
gapminder <- read.csv("https://raw.githubusercontent.com/swcarpentry/r-novice-gapminder/gh-pages/_episodes_rmd/data/gapminder-FiveYearData.csv")

```

```{r}
gapminder %>%
  map_chr(class)
```

`map` operateds on all columns
`dplyr::summarise` summarises specified columns


```{r}
gapminder %>%
  map_df(~data.frame(num = n_distinct(.)))

# summarise a single column
gapminder %>%
  summarise(num = n_distinct(year))

# be careful - this operates on two columns as if they were one - not we want!
gapminder %>%
  summarise(num = n_distinct(c(year, country)))

```

`.id` argument of `map` takes the name of the element being iterated over and returns it:

```{r}
gapminder %>%
  map_df(~data.frame(num = n_distinct(.)), .id = "name")
```

Trick: figure out what the code would be to do what you want for a single element, then paste it into `map_df`:

```{r}
# count distinct values of column
data.frame(num = n_distinct(gapminder$country))

# count distinct values of all columns
map_df(gapminder, ~data.frame(num = n_distinct(.)))
```

Map over two objects using `map2`: `map2(.x = object1, .y = object2, .f = function(.x, .y))`

This works like Python's zip function. ie pairwise elements, not every combination. 

For plotting a scatter of two variables for every country-year combination, we first need a trick for getting every combination:

```{r}
# df of all combinations
cont_year <- gapminder %>% distinct(continent, year)

# extract the variable pairrs as separate vectors
continents <- cont_year %>% pull(continent) %>% as.character
years <- cont_year %>% pull(year)

continents
years

```
Then write a function for a single combination
```{r}
.x <- continents[1]
.y <- years[1]

gapminder %>%
  filter(continent == .x, year == .y) %>%
  ggplot(aes(x = gdpPercap, y = lifeExp)) +
    geom_point() +
    ggtitle(paste(.x, "-", .y))
  
```

And paste it into `map2`:

```{r}
plot_list <- map2(
  .x = continents, 
  .y = years, 
  .f = ~(gapminder %>%
           filter(continent == .x, year == .y) %>%
           ggplot(aes(x = gdpPercap, y = lifeExp)) +
            geom_point() +
            ggtitle(paste(.x, "-", .y))
  ))

plot_list[[1]]

plot_list[[40]]
```

